---
title: "Precipitation in Santa Barbara County and its effect on the Cachuma Reservoir"
author: "Daniel Kerstan"
date: "11/29/2021"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, error = FALSE)
library(modelr)
library(broom)
library(sf)
library(sp)
library(here)
library(lubridate)
library(patchwork)
library(readxl)
library(janitor)
library(raster)
library(gstat)
library(automap)
library(viridis)
library(tidyverse)
library(ggmap)
library(gridExtra)
library(equatiomatic)
```
# Purpose
**The purpose of this document is to show my approach to understanding precipitation in Santa Barbara County by accessing and analyzing precipitation data sets and then applying that analysis to understand its effect on the Cachuma Reservoir.**


# Precipitation

### Gathering Data
Rainfall data was gathered from the [Santa Barbara County Website](https://www.countyofsb.org/pwd/dailyrainfall.sbc).

This data consisted of 81 separate .xls files, each from a different rainfall gauge in Santa Barbara County, which I put into a single folder.

I then created a function to read and clean these files. A for loop was used to apply this function to all 81 files and put all of the data into a single data frame.

A separate function and for loop was used to extract the location data for each rainfall gauge.

After this process I had two data frames, one with daily rainfall totals for each date based on station id and another with the latitude and longitude coordinates of each station (rainfall gauge).

```{r}
# Precipitation data
precip_files <- list.files(path = here("precip"))  
  
read_precip <- function(file) {
  df <- read_excel(here("precip", file), sheet = 1, skip = 9)
  colnames(df) <- c("station_id", "water_year", "year", "month", "day", "daily_rain", "code")
  df <- df %>%   unite(date, c(year, month, day), sep = "-") %>% 
  mutate(date = ymd(date)) %>% 
  clean_names() %>% 
  select(station_id, date, daily_rain)
  return(df)
}

combined <- data.frame()

for (i in precip_files) {
  data <- read_precip(i)
  combined <- rbind(combined, data)
}

# Precipitation locations
meta <- function(file) {
  read_excel(here("precip", file), sheet = 1, skip = 3, n_max = 2) %>% 
  pivot_longer(cols = everything(), names_to = "id", values_to = "coord")
}

combined_meta <- data.frame()

for (i in precip_files) {
  data <- meta(i)
  combined_meta <- rbind(combined_meta, data)
}

combined_meta2 <- combined_meta %>% 
  separate(col = id, into = c("a", "b", "c"), sep = " ") %>% 
  mutate(a = str_replace(a, "#", "")) %>% 
  separate(col = coord, into = c("lat", "lon", "elev"), sep = ",") %>% 
  mutate(lat = str_replace(lat, "Lat ", ""),
         lon = str_replace(lon, "Long ", ""), 
         elev = str_replace(elev, "Elev ", ""),
         elev = str_replace(elev, "ft", "")) %>% 
  rename(id = a) %>% 
  select(id, lat, lon, elev) %>% 
  mutate(id = str_replace(id, " ", ""), lat = str_replace(lat, " ", ""), lon = str_replace(lon, " ", ""), elev = str_replace(elev, " ", "")) %>% 
  separate(lat, into = c("latA", "latB", "latC"), sep = "-") %>% 
  separate(lon, into = c("lonA", "lonB", "lonC"), sep = "-") %>% 
  mutate(latA = as.numeric(latA), latB = as.numeric(latB), latC = as.numeric(latC), lonA = as.numeric(lonA), lonB = as.numeric(lonB), lonC = as.numeric(lonC)) %>% 
  mutate(lat_deg_dec = latA + (latB/60) + (latC/3600), lon_deg_dec = lonA + (lonB/60) + (lonC/3600)) %>% 
  rename(lat = lat_deg_dec, lon = lon_deg_dec) %>% 
  select(id, lat, lon, elev) %>% 
  rename(station_id = id) %>% 
  mutate(station_id = as.numeric(station_id)) %>% 
  mutate(elev = as.numeric(elev)) %>% 
  mutate(lon = lon*-1)

# finalized data frames
precip <- combined %>% 
  drop_na()

meta <- combined_meta2
precip_withcoord <- left_join(precip, meta, by = "station_id")
```

### Wrangling the Data

The next step was to wrangle this data to produce data frames which could be used for analysis.

```{r}
# daily precip averaged across the entire county
precip_avg <- precip %>% 
  group_by(date) %>% 
  summarize(avg_rain = mean(daily_rain)) %>% 
  mutate(year = year(date))

# precip yearly totals based on station id
precip_year_all <- precip %>% 
  mutate(year = year(date)) %>% 
  group_by(station_id, year) %>% 
  summarize(year_precip = sum(daily_rain))

# precip yearly totals averaged over entire county
precip_year_avg <- precip_year_all %>% 
  ungroup() %>% 
  group_by(year) %>% 
  summarize(avg_precip = mean(year_precip))

# precip monthly totals based on station id
precip_month_all <- precip %>% 
  mutate(month = month(date), year = year(date)) %>% 
  unite(monthyear, c(month, year), sep = "-") %>% 
  mutate(monthyear = my(monthyear)) %>% 
  group_by(station_id, monthyear) %>% 
  summarize(month_precip = sum(daily_rain)) %>% 
  mutate(year = year(monthyear))

# precip monthly totals averaged over entire county
precip_month_avg <- precip_month_all %>% 
  ungroup() %>% 
  group_by(monthyear) %>% 
  summarize(avg_precip = mean(month_precip)) %>% 
  mutate(year = year(monthyear))
```

### Plotting the points
I first used `ggmap::get_stamenmap()` to get a map of Santa Barbara County.

I then created another data frame with yearly rainfall totals averaged over all the years in the data set based on station id. 

This allowed me to plot these points onto the map.

```{r, fig.cap= "Average yearly precipitation in Santa Barbara County"}
### get a basemap
sb_map <- get_stamenmap(bbox = c(left = -120.65,
                                    bottom = 34.39694,
                                    right = -119.4622,
                                    top = 35.07694),
          maptype = "terrain", 
          crop = TRUE,
          zoom = 10)

# get yearly precipitation average for all years for each station
precip_year_plot <- precip_year_all %>% 
  ungroup() %>% 
  group_by(year, station_id) %>% 
  summarize(year_precip) %>% 
  left_join(meta, by = "station_id") %>% 
  ungroup() %>% 
  group_by(station_id, lat, lon) %>% 
  summarize(year_precip = mean(year_precip))

#create a plot of points on a map of Santa Barbara County
ggmap(sb_map, darken = .2) +
  geom_point(data = precip_year_plot, mapping = aes(x = lon, y = lat, color = year_precip), size = 2) +
  scale_color_gradient(low = "lightblue", high = "blue4") +
  labs(title = "Yearly Precipitation in Santa Barbara County", color = "Yearly Precipitation (in)") +
  theme(legend.position = "bottom")
```

### Kriging

The next step was to spatially interpolate the data to get a better idea of the precipitation in the entire county.

I decided to use the form of spatial interpolation known as Ordinary Kriging.

The first step in this process was to create a variogram, which describes the spatial dependence. This was done using the `gstat::variogram()`.
The function `automap::autofitVariogram()` was used to choose the model that best fits the data.
Next, after defining a target grid, the `gstat::krige()` function was used to generate the set of predictions.

```{r, fig.cap= "Spatial interpolation of average yearly precipitation in Santa Barbara County", results='hide'}
precip_sf <- st_as_sf(precip_year_plot, coords = c("lon", "lat"), crs = 4326) %>% 
  cbind(st_coordinates(.))

v_emp_OK <- gstat::variogram(
  year_precip~1,
  as(precip_sf, "Spatial"))

v_mod_full <- automap::autofitVariogram(year_precip~1, as(precip_sf, "Spatial"))

v_mod <- v_mod_full$var_model

grd_sf <- precip_sf %>% 
  st_bbox() %>% 
  st_as_sfc() %>% 
  st_make_grid( 
  cellsize = c(.005, .005), 
  what = "centers"
  ) %>%
  st_as_sf() %>% 
  cbind(., st_coordinates(.))

grd_sp <- as(grd_sf, "Spatial") 
gridded(grd_sp) <- TRUE             
grd_sp <- as(grd_sp, "SpatialPixels") 

OK <- krige(
  year_precip~1,                      
  as(precip_sf, "Spatial"), 
  grd_sp,                
  model = v_mod         
  )

df <- rasterToPoints(raster(OK)) %>% 
  as_tibble()
colnames(df) <- c("X", "Y", "Z")

OK_plot <- ggmap(sb_map, darken = .2) +
  geom_raster(data = df, aes(x = X, y = Y, fill = Z)) +
  coord_cartesian() +
  scale_fill_gradient2(low = "transparent", mid = "blue4", high = "blue4", midpoint = 30) +
  labs(title = "Yearly Precipitation in Santa Barbara County", fill = "Yearly Precipitation (in)") +
  theme(legend.position = "bottom")

OK_plot
```
  

    
To get a better understanding of how precipitation has been changing over the years, I looped this process over a subset of years.

```{r, fig.cap= "Spatially interpolated yearly precipitation in Santa Barbara County", results='hide'}
krig_precip <- function(desired_year) {
  precip_year_plot <- precip_year_all %>% 
  ungroup() %>% 
  group_by(year, station_id) %>% 
  summarize(year_precip) %>% 
  left_join(meta, by = "station_id") %>% 
  filter(year == desired_year)
  
  precip_sf <- st_as_sf(precip_year_plot, coords = c("lon", "lat"), crs = 4326) %>% 
  cbind(st_coordinates(.))

v_emp_OK <- gstat::variogram(
  year_precip~1,
  as(precip_sf, "Spatial"))

v_mod_full <- automap::autofitVariogram(year_precip~1, as(precip_sf, "Spatial"))

v_mod <- v_mod_full$var_model

grd_sf <- precip_sf %>% 
  st_bbox() %>% 
  st_as_sfc() %>% 
  st_make_grid( 
  cellsize = c(.005, .005), 
  what = "centers"
  ) %>%
  st_as_sf() %>% 
  cbind(., st_coordinates(.))

grd_sp <- as(grd_sf, "Spatial") 
gridded(grd_sp) <- TRUE             
grd_sp <- as(grd_sp, "SpatialPixels") 

OK <- krige(
  year_precip~1,                      
  as(precip_sf, "Spatial"), 
  grd_sp,                
  model = v_mod         
  )

df <- rasterToPoints(raster(OK)) %>% 
  as_tibble()
colnames(df) <- c("X", "Y", "Z")

OK_plot <- ggmap(sb_map, darken = .2) +
  geom_raster(data = df, aes(x = X, y = Y, fill = Z)) +
  coord_cartesian() +
  scale_fill_gradient2(low = "transparent", mid = "blue4", high = "black", midpoint = 30, limits = c(0,55)) +
  labs(title = paste(desired_year), fill = "in_rain") +
  theme(axis.title.x = element_blank(), axis.title.y = element_blank(), axis.text.x = element_blank(), axis.text.y = element_blank(), legend.position = "bottom", legend.title = element_blank())

return(OK_plot)
}

for (i in 2015:2021) {
  plot <- krig_precip(i)
  assign(paste0("plot_", i), plot)
}

plot_county <- ggplot(filter(precip_year_avg, year >= 2015), aes(x = year, y = avg_precip)) +
  geom_col(color = "blue4", fill = "blue4") +
  labs(title = "Countywide Average Rain (in)", x = "Year") +
  theme(axis.title.y = element_blank(), plot.title = element_text(size = 8))
  


grid.arrange(plot_2015, plot_2016, plot_2017, plot_2018, plot_2019, plot_2020, plot_2021, plot_county, nrow = 2)


```

# Cachuma Reservoir

With the precipitation data sorted out, I could now apply it to understanding the effect of precipitation on water level change in the Cachuma Reservoir.
The Cachuma Reservoir is heavily relied upon by the city of Santa Barbara, which is entitled to 32.19% of its available water. Understanding how rainfall affects this reservoir is vitally important.


## Reservoir data

I found reservoir level data for the Cachuma Reservoir on the [County of Santa Barbara Public Works website](https://rain.cosbpw.net/sensor/?time_zone=US%2FPacific&site_id=105&site=70729dd9-97d4-430a-9271-7b6c195b49be&device_id=1&device=5d7a3129-708d-4881-9886-f84c6686ab41&data_start=2012-10-29%2000%3A00%3A00&data_end=2012-11-28%2023%3A59%3A59&bin=3600&range=Custom%20Range&markers=false&legend=true&thresholds=true&refresh=off&show_raw=true&show_quality=true).

Fairly consistent data was provided going back to the year 2015. Reservoir level were measured at 15 minute intervals in the units of feet.

```{r}
## Cachuma data read in

url1 <- "https://rain.cosbpw.net/export/file/?site_id=105&site=70729dd9-97d4-430a-9271-7b6c195b49be&device_id=1&device=5d7a3129-708d-4881-9886-f84c6686ab41&mode=&hours=&data_start="

url2 <- "-01-01%2000:00:00&data_end="

url3 <- "-12-31%2023:59:59&tz=US%2FPacific&format_datetime=%25Y-%25m-%25d+%25H%3A%25i%3A%25S&mime=txt&delimiter=tab"

read_cachuma <- function(year) {
  df <- read_delim(paste0(url1, year, url2, year, url3), delim = "\t")
  df <- df %>% mutate(Reading = ymd_hms(Reading)) %>% 
  mutate(date = lubridate::as_date(Reading)) %>% 
  group_by(date) %>% 
  summarize(level = mean(Value)) %>% 
  mutate(year = year(date))
  return(df)
}

years_cachuma <- c("2021", "2020", "2019", "2018", "2017", "2016", "2015", "2013", "2012")

cachuma <- data.frame()

for (i in years_cachuma) {
  data <- read_cachuma(i)
  cachuma <- rbind(cachuma, data)
}

cachuma <- cachuma %>% 
  arrange(date)

cachuma$level_change = NA

for (i in 2:length(cachuma$date)) {
  if ((cachuma$date[i] - cachuma$date[i-1]) == 1) {
    cachuma$level_change[i] <- cachuma$level[i] - cachuma$level[i-1]
  }
}

### cachuma daily
cachuma_precip <- cachuma %>% 
  full_join(precip_avg) %>% 
  filter(year >= 2015) %>% 
  arrange(date)

cachuma_precip_all <- cachuma %>% 
  full_join(precip) %>% 
  filter(year >= 2015) %>% 
  arrange(date)

### cachuma monthly
cachuma_precip_month <- cachuma %>% 
  mutate(month = month(date)) %>% 
  unite(monthyear, c(month, year), sep = "-") %>% 
  mutate(monthyear = my(monthyear)) %>% 
  group_by(monthyear) %>% 
  summarize(monthly_level_change = sum(level_change)) %>% 
  full_join(precip_month_avg) %>% 
  filter(year >= 2015) %>% 
  arrange(monthyear)

### cachuma monthly all
cachuma_precip_month_all <- cachuma %>% 
  mutate(month = month(date)) %>% 
  unite(monthyear, c(month, year), sep = "-") %>% 
  mutate(monthyear = my(monthyear)) %>% 
  group_by(monthyear) %>% 
  summarize(monthly_level_change = sum(level_change)) %>% 
  full_join(precip_month_all) %>% 
  filter(year >= 2015) %>% 
  arrange(monthyear) 

```

  
    
With this data, I was able to make some quick plots.

```{r, fig.cap= "Cachuma Reservoir levels and level changes over time"}
cachuma_quick_1 <- ggplot(cachuma_precip, aes(x = date)) +
  geom_line(aes(y = level)) +
  scale_y_continuous(limits = c(600, 800)) +
  labs(title = "Reservoir Level (ft)", x = "Date") +
  theme(axis.title.y = element_blank(), axis.title.x = element_blank(), plot.title = element_text(size = 9))

cachuma_quick_2 <- ggplot(cachuma_precip, aes(x = date)) +
  geom_line(aes(y = level_change)) +
  labs(title = "Change in Reservoir Level (ft)", x = "Date")+
  theme(axis.title.y = element_blank(), axis.title.x = element_blank(), plot.title = element_text(size = 9))


cachuma_quick_3 <- ggplot(cachuma_precip_month, aes(x = monthyear, y = monthly_level_change)) +
  geom_col() +
  labs(title = "Monthly Change in Reservoir Level (ft)", x = "Date")+
  theme(axis.title.y = element_blank(), axis.title.x = element_blank(), plot.title = element_text(size = 9))


cachuma_quick_1 / cachuma_quick_2 / cachuma_quick_3


```

  
# Statistical Analysis

### More Kriging

After matching up the reservoir data with the precipitation data, I wanted to see which stations recorded rainfall levels that correlated most with changes in reservoir water level. To do this, I created a function that assigns an r squared value to a station based on the results of a regression analysis of the effect of monthly total rainfall at that station and the monthly change in reservoir level. 

Initially I performed a simple linear regression using the `lm()` function. Later on in the analysis process, I found that a polynomial regression, done using `lm(x ~ poly(y, 2))`, was able to fit the data much better. This is why I used a polynomial regression model for this function. The equation for this model is $$\operatorname{monthly\_level\_change} = \alpha + \beta_{1}(\operatorname{month\_precip}) + \beta_{2}(\operatorname{month\_precip^2}) + \epsilon$$  

I then looped the function over all of the 81 stations and put the results into a new data frame.



```{r}
station <- function(station) {
  station_eval <- cachuma_precip_month_all %>% 
    filter(station_id == station)
  model <- lm(monthly_level_change~poly(month_precip,2), data = station_eval)
  sum_model <- summary(model)
  r_sq <- sum_model$r.squared
  return(r_sq)
}

stations <- unique(cachuma_precip_month_all$station_id)

station_df <- data.frame(station_id = stations, r_sq = NA) %>% 
  arrange(station_id)

for (i in 1: length(station_df$station_id)) {
  station_df$r_sq[i] <- station(station_df$station_id[i])
}

station_df <- station_df %>% 
  left_join(meta)

station_df <- select(station_df, -elev)

station_df_most <- station_df %>% 
  arrange(-r_sq)

station_df_most <- head(station_df_most, 1)

```


With these r squared values, I thought a good way to visualize this would be with another spatial interpolation. This interpolation would show differences in correlation between rainfall and reservoir water level change across the county. 


```{r, fig.cap= "Correlation between precipitation and reservoir level change based on r squared value attained through linear regression analysis", results='hide'}
correlation_sf <- st_as_sf(station_df, coords = c("lon", "lat"), crs = 4326) %>% 
  cbind(st_coordinates(.))

v_emp_OK <- gstat::variogram(
  r_sq~1,
  as(correlation_sf, "Spatial"))

v_mod_full <- automap::autofitVariogram(r_sq~1, as(correlation_sf, "Spatial"))

v_mod <- v_mod_full$var_model

grd_sf <- correlation_sf %>% 
  st_bbox() %>% 
  st_as_sfc() %>% 
  st_make_grid( 
  cellsize = c(.005, .005), 
  what = "centers"
  ) %>%
  st_as_sf() %>% 
  cbind(., st_coordinates(.)) 

grd_sp <- as(grd_sf, "Spatial") 
gridded(grd_sp) <- TRUE             
grd_sp <- as(grd_sp, "SpatialPixels")

OK <- krige(
  r_sq~1,                      
  as(correlation_sf, "Spatial"), 
  grd_sp,                
  model = v_mod         
  )

df <- rasterToPoints(raster(OK)) %>% 
  as_tibble()
colnames(df) <- c("X", "Y", "Z")

OK_plot <- ggmap(sb_map, darken = .2) +
  geom_raster(data = df, aes(x = X, y = Y, fill = Z)) +
  coord_cartesian() +
  scale_fill_gradient2(low = "transparent", mid = "transparent", high = "red", midpoint = 0.55) +
  theme(legend.position = "bottom")+
  geom_text(data = station_df, inherit.aes = FALSE, aes(x = lon, y = lat, label = station_id), size = 3) +
  labs(title = "Correlation with Cachuma Reservoir Level", fill = "R Squared Value") +
  theme(axis.title.x = element_blank(), axis.title.y = element_blank())

OK_plot
```

The station with the highest correlation was `r station_df_most$station_id` with an r squared value of `r round(station_df_most$r_sq, 3)`

Plotting it in this way was a good way to check and make sure that this correlation makes sense. Station 238 does appear to be in an area that would drain into the Cachuma Reservoir.  

I chose to use station 238 for the remainder of my analysis.
```{r}
station_238_monthly <- cachuma_precip_month_all %>% 
  filter(station_id == 238)

station_238_daily <- cachuma_precip_all %>% 
  filter(station_id == 238)
```


Here is the summary of the linear regression model for station 238.
```{r}
model <- lm(monthly_level_change~ poly(month_precip,2), data = station_238_monthly)
model_summary <- summary(model)
model_summary

```

I then put together a graph of the data with the linear model, as well as graphs of the residuals.


```{r, fig.cap= "Relationship between precipitation at station 238 and reservoir water level change"}
predictions <- station_238_monthly %>% 
  add_predictions(model) %>% 
  mutate(residuals = monthly_level_change - pred)

model_plot <- ggplot(station_238_monthly, aes(x = month_precip, y = monthly_level_change)) +
  geom_point() +
  geom_line(data = predictions, aes(x = month_precip, y = pred), color = "blue") +
  labs(x = "Monthly Precipitation", y = "Monthly Change in Reservoir Level") +
  theme(axis.title.y = element_text(size = 8), axis.title.x = element_text(size = 10))


res_hist <- ggplot(predictions) +
  geom_histogram(aes(residuals))

res_point <- ggplot(predictions) +
  geom_point(aes(x = month_precip, y = residuals))

model_plot / (res_hist + res_point)

```

# Plots

Revisiting the plots I made of the Cachuma Reservoir water levels earlier, I could now confidently add the precipitation data from station 238.

```{r, fig.cap= "Daily and monthly interaction between precipitation at station 238 and reservoir water level"}
final_daily_plot <- ggplot(station_238_daily, aes(x = date)) +
  geom_rect(aes(ymin=600, ymax = 800, xmin = date-7, xmax=date+7, fill = daily_rain)) +
  scale_fill_gradient(low = "transparent", high = "royalblue1", na.value = "transparent") +
  geom_line(aes(y = level)) +
  scale_y_continuous(limits = c(600, 800)) +
  labs(y = "Reservoir Level (ft)", fill = "Daily Rain (in)") +
  theme(axis.title.x = element_blank(), axis.title.y = element_text(size = 9), legend.title = element_text(size = 8))

final_daily_change_plot <- ggplot(station_238_daily, aes(x = date)) +
  geom_rect(aes(ymin=-5, ymax = 20, xmin = date-7, xmax=date+7, fill = daily_rain)) +
  scale_fill_gradient(low = "transparent", high = "royalblue1", na.value = "transparent") +
  geom_line(aes(y = level_change)) +
  labs(y = "Level Change (ft)", fill = "Daily Rain") +
  theme(axis.title.x = element_blank(), axis.title.y = element_text(size = 9), legend.title = element_text(size = 8))
  


final_monthly_plot <- ggplot(station_238_monthly) +
  geom_rect(aes(ymin=-10, ymax = 50, xmin = monthyear, xmax=monthyear+30, fill = month_precip), alpha = 0.5) +
  scale_fill_gradient2(low = "transparent", mid = "royalblue1", high = "darkblue", midpoint = 5, na.value = "transparent") +
  geom_col(aes(y = monthly_level_change, x = monthyear + 15)) +
  labs(y = "Level Change (ft)", fill = "Monthly Rain") +
  theme(axis.title.x = element_blank(), axis.title.y = element_text(size = 9), legend.title = element_text(size = 8))


final_daily_plot / (final_daily_change_plot + final_monthly_plot)
```


## References

“SB County Public Works Water Resources Hydrology - Daily Rainfall XLS.” n.d. Accessed November 30, 2021. https://www.countyofsb.org/pwd/dailyrainfall.sbc.  

“Sensor.” n.d. Accessed November 30, 2021. https://rain.cosbpw.net/sensor/?time_zone=US%2FPacific&site_id=105&site=70729dd9-97d4-430a-9271-7b6c195b49be&device_id=1&device=5d7a3129-708d-4881-9886-f84c6686ab41&data_start=2012-10-29%2000%3A00%3A00&data_end=2012-11-28%2023%3A59%3A59&bin=3600&range=Custom%20Range&markers=false&legend=true&thresholds=true&refresh=off&show_raw=true&show_quality=true.












